\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-04-20 doc/incertae_sedis.tex
minor

Note that, for the purposes of the supertree method, all of the
information from the taxonomic tree is contained in the set of rooted
splits.
Thus, one could imagine replacing that taxonomy,
$\taxonomy_{p}$, with a set of trees $\mathcal{R}(\taxonomy_{p})$ each
of which has a single internal edge such that each edge in
$\taxonomy_{p}$is included in one tree in
$\mathcal{R}(\taxonomy_{p})$.
If the order of trees is set to be (with
some abuse of notation):
$\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p}) + 
\mathcal{Y}(\taxonomy_{p})$, then using $\mathcal{R}(\taxonomy_{p})$ in
place of $\taxonomy_{p}$in the supertree algorithm would produce the
same output.

We define rooted splits for each edge by noting that removing the edge
would produce two trees.
The leaf label sets of these trees the
partition the full leaf set $\mathcal{L}$ into two groups.
The include
set $\mathcal{I}(e)$ which does not contain the root, and the exclude
set $\mathcal{E}(e)$ which does contain the root.
Such a split may be
written \[ \mathcal{I}(e)|\bullet\mathcal{E}(e).\] where the
$\bullet$ indicates the root.
The include set is the set of leaves
that are descendants of the daughter node of the edge.
The exclude set
is the set of leaves that are not descendants of that node.
If no taxa
are \emph{incertae sedis}, then the exclude set for a node is just the
total tip set minus the include set for the node:
$\mathcal{E}(e)  =\mathcal{L}-\mathcal{I}(e)$.

For a node $n$ on the tip-ward side of an edge $e$, we may also write
$\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for
$\mathcal{E}(e)$.
For any two rooted splits $A=A_{1}|\bullet A_{2}$
and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$ if
$B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.

@@

\paragraph{Information content of the emph{incertae sedis }annotation}

In order to incorporate \emph{incertae sedis }taxa into a supertree
analyses, we must arrive at an operational definition of the meaning
of the \emph{incertae sedis} label.
The core meaning of the phrase is
that the annotated taxon may actually be a member of one of the taxa
that appear as siblings to it in the heirarchical representation of
the taxonomy.
In the most expansive interpretation, the author of the
taxonomy is stating that the \emph{incertae sedis} taxon could be
placed anywhere inside one of the sibling taxa or their descendants.
However, frequently only a subset of the possible placement points
would be viewed as plausible.
Ideally, the \emph{incertae sedis}
annotation some taxon $x$ would be accompanied by a specific
statement about which other taxa might include $x$.
Detailed
readings of the character argumentation used by taxonomists could
provide this sort of specific information in many cases, but we lack
such information on the scale of the entire OTT.
Here we consider a
few operational approaches for handling the \emph{incertae sedis}
label.

In terms of rooted split interpretation, an \emph{incertae sedis}
taxon differs from a normal taxon because its leaves should not
automatically be mentioned in the exclude set of all of the edges that
does not descend from.
If a taxon rooted at node $n$ is marked as
\emph{incertae sedis}, then the leaf labels that are descendants of
$n$ will still be in the include set for $n$ and all its ancestors,
but these labels will not necessarily be the exclude set for all other
nodes/edges.
They may simply be omitted in some rooted split
representations, resulting in partial splits.

@@

The most straightforward approach for increasing the taxonomic
completeness of the supertree is simply to no longer prune taxa from
$\taxonomy$ because they are flagged as \emph{incertae sedis}, and
then run the rest of the pipeline of \citet{redelings2017supertree}
unaltered.
However this approach results in a less resolved tree
because taxonomic vagueness (in the form of the \emph{incertae sedis}
annotation) is interpreted as an unwarranted statement that the
\emph{incertae sedis} taxa are disallowed from being members of their
sibling taxa.

>>>>

Note that, for the purposes of the supertree method, all of the
information from the taxonomic tree is contained in the set of rooted
splits.
Thus, one could imagine replacing that taxonomy,
$\taxonomy_{p}$, with a set of rooted splits $\mathcal{R}(\taxonomy_{p})$ 
such that each edge in
$\taxonomy_{p}$ is encoded in one split in
$\mathcal{R}(\taxonomy_{p})$.
If the order of trees is set to be:
$\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p}) + 
\mathcal{Y}(\taxonomy_{p})$, then using $\mathcal{R}(\taxonomy_{p})$ in
place of $\taxonomy_{p}$ in the supertree algorithm would produce the
same output.

As is typically done, we define rooted splits for each edge by
noting that removing the edge
would produce two trees.
The leaf label sets of these trees the
partition the full leaf set $\mathcal{L}$ into two groups.
The include
set $\mathcal{I}(e)$ which does not contain the root, and the exclude
set $\mathcal{E}(e)$ which does contain the root.
Such a split may be
written \[ \mathcal{I}(e)|\bullet\mathcal{E}(e).\] where the
$\bullet$ indicates the root.
The include set is the set of leaves
that are descendants of the daughter node of the edge.
The exclude set
is the set of leaves that are not descendants of that node.
If no taxa
are \emph{incertae sedis}, then the exclude set for a node is just the
total tip set minus the include set for the node:
$\mathcal{E}(e)  =\mathcal{L}-\mathcal{I}(e)$.
For a node $n$ on the tip-ward side of an edge $e$, we may also write
$\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for
$\mathcal{E}(e)$.
For any two rooted splits $A=A_{1}|\bullet A_{2}$
and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$ if
$B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.

@@

\paragraph{Information content of the \emph{incertae sedis }annotation}

In order to incorporate \emph{incertae sedis }taxa into a supertree
analyses, we must arrive at an operational definition of the meaning
of the \emph{incertae sedis} label.
The core meaning of the phrase is
that the annotated taxon may actually be a member of one of the taxa
that appear as siblings to it in the heirarchical representation of
the taxonomy.
In the most expansive interpretation, the author of the
taxonomy is stating that the \emph{incertae sedis} taxon could be
placed anywhere inside one of the sibling taxa or their descendants.
However, frequently only a subset of the possible placement points
would be viewed as plausible.

Ideally, the \emph{incertae sedis}
annotation some taxon $x$ would be accompanied by a specific
statement about which other taxa might include $x$.
Detailed
readings of the character argumentation used by taxonomists could
provide this sort of specific information in many cases, but we lack
such information on the scale of the entire OTT.
Here we consider a
few operational approaches for handling the \emph{incertae sedis}
label.

In terms of rooted split interpretation, an \emph{incertae sedis}
taxon differs from a normal taxon because its leaves should not
automatically be mentioned in the exclude set of all of the edges that
does not descend from.
If a taxon rooted at node $n$ is marked as
\emph{incertae sedis}, then the leaf labels that are descendants of
$n$ will still be in the include set for $n$ and all its ancestors,
but these labels will not necessarily be the exclude set for all other
nodes/edges.
They may simply be omitted in some rooted split
representations, resulting in ``partial splits'' -- those which
do not necessarily mention all of the taxa.
Thus, when interpreted correctly the \emph{incertae sedis} annotations
are reducing the amount of information that the taxonomy brings to
the supertree problem.

@@

The most straightforward approach for increasing the taxonomic
completeness of the supertree is simply to no longer prune taxa from
$\taxonomy$ because they are flagged as \emph{incertae sedis}, and
then run the rest of the pipeline of \citet{redelings2017supertree}
unaltered.
However this approach results in a less resolved tree
because taxonomic vagueness (in the form of the \emph{incertae sedis}
annotation) is interpreted as an unwarranted statement that the
\emph{incertae sedis} taxa are disallowed from being members of their
sibling taxa.
In essence, this interpretation would overinterpret the taxonomic
    hierarchy by treating all of the groupings as statements about
    full rooted splits, when some should be viewed as partial splits.
